---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(simpletrie)
```

# simpletrie

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
![](https://img.shields.io/badge/dependencies-zero-blue.svg)
<!-- badges: end -->

`simpletrie` is a simple [trie](https://en.wikipedia.org/wiki/Trie) implementation
in plain R code.

A trie is a structure for representing certain data as a tree of nodes. This 
data representation allows for fast queries of certain types.

`simpletrie` can build a trie from 175k words in about 3s.


## What's in the box

* `trie_create()` - create a trie, and optionally initialise with a dictionary of words
* `trie_insert()` - insert words into a trie
* `trie_find_subwords()` - find all sub-words in the trie using the characters in the given word


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/simpletrie) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/simpletrie)
```

## Example: Finding sub-words from a given word

My main use-case for `simpletrie` is for looking up scrabble words i.e. to find all possible sub-words of a given word with:

* the characters in any order
* being able to specify a *wildcard* to indicate that any letter can be used
  i.e. the `.` character


```{r example}
library(simpletrie)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Read the ENABLE word list
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
words <- tolower(readLines("working/enable1.txt"))
length(words)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create trie strcture
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
system.time({
  trie <- trie_create(words)
})

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Find all words which can be made with 'hi.rstats'
# '.' is a wildcard, and can match any letter. (Like a blank tile in scrabble)
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
system.time({
  subwords <- trie_find_subwords(trie, 'hirstats.')
})

length(subwords)
head(subwords, 100)
```


## What's a trie?

A *trie* is an ordered data structure represented as a tree of nodes.

The image below shows a very small trie which represents the words: 'cat', 'cog', 'dog' and 'dote.'

This representation allows for very fast queries of certain types.

```{dot echo=FALSE, out.width = "50%"}
digraph D {
  root [label="root node"]
  c
  a 
  t
  o1 [label="o"]
  g1 [label="g"]
  d1 [label="d"]
  o2 [label="o"]
  g2 [label="g"]
  t3 [label="t"]
  e4 [label="e"]
  
  root -> c;
  root -> d1;
  
  c -> a -> t;
  c -> o1 -> g1;
  d1 -> o2 -> g2;
  o2 -> t3 -> e4;
}
```


## Technical bits

The trie is implemented here as nested environments.  Environments are
created with `new.env(parent = emptyenv(), hash = FALSE)`.

Using the `emptyenv()` as parent saves some memory, as the parent
environment is never needed in this process.

`hash` is set to `FALSE` as using a hash table expands the memory
usage by quite a bit - even if `size` is set to limit the initial hash
table size.

Using a hash table on the environment doesn't 
appreciably change the speed of trie creation or preforming look-ups.



## Related Software + Resources

* [Trie on Wikipedia](https://en.wikipedia.org/wiki/Trie)
* [iptrie](https://github.com/hrbrmstr/iptrie) - Efficiently Store and Query ‘IPv4’ Internet Addresses with Associated Data
* [triebeard](https://cran.r-project.org/package=triebeard)
* [rtrie](https://cran.r-project.org/web/packages/rtrie/index.html) - was previously on cran but has now been archived
* [AhoCorasickTrie](https://cran.r-project.org/web/packages/AhoCorasickTrie/index.html)
* Some raw wordlists e.g. ENABLE1 and SOWPODS can be found [here](http://norvig.com/ngrams/)
* [Bare bones trie creation in 10 lines of R](http://stackoverflow.com/questions/27060453/how-to-build-an-alphabetical-tree-from-a-list-of-words-in-r) 

## Acknowledgements

* R Core for developing and maintaining such a wonderful language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
